LabMate: supporting types for MATLAB

### 2024-03-28 writing/IMEKOWC2024/LabMate/paper.tex
[ writing ] draft intro and conclusions

\title{LabMate: supporting types for MATLAB}
\author{Conor McBride, Georgi Nakov, Fredrik Nordvall Forsberg, Neil Ghani, Alasdair Forbes, Keith Lines, Ian Smith}
\address{University of Strathclyde, National Physical Laboratory}
%TODO: Affiliations

@@

% types help

% at low cost  (ie: don't need to start from scratch)

% Matlab popular etc

@@

\section{Conclusions}

% future work etc

>>>>

\title{LabMate: supporting types for MATLAB}
\author{Neil Ghani, Conor McBride, Georgi Nakov, Fredrik Nordvall Forsberg, Alasdair Forbes, Keith Lines, Ian Smith}
\address{University of Strathclyde, National Physical Laboratory}
%TODO: Affiliations

@@

MATLAB is a key workhorse in many scientific and engineering
disciplines that are heavily reliant on numerical methods. It helps us
do powerful things. However, as with all software, MATLAB code may
contain implementation errors and bugs. In good programming practice
for MATLAB developers, programmers use comments to make clear what
physical systems their data concern and how the data should be
interpreted, specifying, e.g., units of measure for
quantities. Regrettably, none of this rich and often disciplined
metadata is perceptible to MATLAB, which instead enforces the
compatibility of producers and consumers of data by run-time checking
of tags which indicate only machine representation, not any form of
meaning. In reality, the programmers document meaning for each other's
benefit but keep the machine in the dark.

To rectify this situation, we have developed LabMate, which is a tool
to reify current virtuous engineering practices as a formal language
of MATLAB comments. MATLAB remains in the dark, but LabMate reads,
assesses, and transforms MATLAB programs in accordance with these
comments.
%
Behind the scenes, LabMate is retrofitting an expressive type system
to MATLAB, with the more of the meaning of programs recorded in their
types.
%
This gives a lightweight and low-cost way for MATLAB programmers to
express their intent, and in a language they are already working in,
rather than starting over from scratch.
%
While type systems and their theory is a well established discipline
of computer science, the development of LabMate has a several novel
advances novel focus on the algebraic structure required to classify
matrices and the meanings of the quantities therein, e.g., their
physical dimensions. LabMate brings advanced type systems to
effectiveness within, rather than instead of, existing scientific and
engineering toolchains and practices.

@@

\section{Conclusions and future work}

We have demonstrated how LabMate can help write MATLAB programs that are correct both in the sense that they avoid runtime errors, as well as dimensional inconsistency for physical dimensions.
%
Getting an error for types not matching can sometimes feel like a stick, but there are also plenty of opportunities for carrots, for example where the specification encoded in the type of the program admits a single unique solution, which could be generated automatically by LabMate.
%
Similarly, LabMate already has support for generating MATLAB code for serialising and deserialising complex data formats to and from files, based on a more high-level description of the data format (for example also taking units of measure into account)~\cite{mgen}.

So far, our focus has been on matrices, as this is what makes MATLAB
unique as a programming language.
%
In the future, we would like to also extend our coverage to
conditionals and loops, which might necessitate some approximation of
what is known about program fragments at runtime.
